---
title: "DataScience-Present-Package"
output: html_document
date: "2023-12-16"
---

Overview
adegenet is a package for a variey of population genetics analyses
-it has a variety of dependencies and recommended packages that can work along with it. THese include:
ape (various phylogenetic tools)
pegas (various population genetic analyses tools)
seqinr (for sequence retrieval and analyses)
hierfstat (for calculation of F-statistics and other population information)
tidyverse}
```{r}
install.packages("ape")
install.packages("pegas")
install.packages("seqinr")
install.packages("hierfstat")
library(ape)
library(pegas)
library(seqinr)
library(hierfstat)
library(tidyverse)

```

adegenet supports 4 types of genetic data files
-genind, genpop, SNPbin, and genlight
-SNPbin and genlight are for large, genomic datasets which are condensed so that analyses can be run on standard laptops

upload data similarly to other data (e.g. read.genepop()), though if you want to convert a .csv you will need to use a special function

```{r}
amac -> read.genepop("Amac-Redo-Genepop.gen", ncode=3)
#this was the name of one of my microsat files, replace with the title of your genetic data file and whatever object name you would like
```


Once your data has been uploaded, you can utilize a variety of statistics, and it has a fairly detailed tutorial. Here are a few examples:

```{r}
hw.test(amac, B=0) 
#testing for Hardy-Weinberg Equilibrium at all loci

wc(amac) 
#gives overall FST and FIS as Weir and Cockerham's unbiased estimators

Fst(as.loci(amac)) 
#FST, FIS, and FIT per locus

amac2ci -> genind2hierfstat(amac)
boot.vc(amac2ci[1], amac2ci[-1])$ci 
#confidence intervals for the overall statistics

pairwise.WCfst(amac)
genet.dist(amac) 
# for pairwise FST or other genetic distance measures, confidence intervals with boot.ppfst()
```

-The defining feature of agegenet is the DAPC (discriminant analysis of principal components)
-first, you need to determine the number of principal components and genetic clusters
  -select the number of PCs that maximize variation while minimizing PCs, usually around where it starts to plateau
  -for clusters, find the minimum BIC value, usually a sharp peak
-then run the DAPC 

```{r}
find.clusters(amac)
#will give 2 prompts
  # choose the # of PCs to retain (>=1)
  # choose the number of clusters (>=2)

dapc(amac)
#to run the dapc
#will also give two prompts
  # choose the # of PCs to retain (>=1)
  # choose the number of discriminant functions to retain

dapcamac -> dapc(amac)
scatter(dapcamac)
#creates the scatter plot. This is customizable, see tutorial for details, several examples below

scatter(dapcamac, scree.da=TRUE, posi.da= "bottomleft")
# moves the DA eigenvalues legend to the bottom left

scatter(dapcamac, scree.da=TRUE, bg="white", pch=14:22, cstar=0, clab=0, col=mycol, leg=TRUE, txt.leg=paste("cluster", 1:9), posi.leg="topleft", posi.da= "bottomleft")
# changes the shape of the data points, changes the colors of the points to a preset (mycol, this can be whetever suite of colors you want), adds a legend with clusters called cluster 1, cluster 2, etc.put in the top left.

#see tutorial for other options. There is a lot available
```

The package has a lot of other features. It also has functions for PCAs, sPCAs, isolation by distance tests (e.g. Mantel). See tutorial for details. 
