---
title: "ephelps_ggpubr.Rmd"
author: "Erin Phelps"
date: "2023-12-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ggpubr
 
ggplot2 package is great for data visualization. However, it can have confusing syntax for beginners in R. Many researchers/scientists use R just for data visualization to produce highly-customizable charts on demand. So, ggpubr makes it easier to produce publication-ready plots using ggplot2.

ggpubr makes it easier to: 
 - add p-values and significance levels
 - Arrange and annotate plots on the same page
 - Change graphical parameters -> colors, labels, etc.


## loading in ggpubr package
Since ggpubr is a wrapper, ggplot also needs to be loaded in.

```{r, message=FALSE}
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(palmerpenguins)
```

## Read in data to make plot


```{r, message=FALSE}
penguins <- penguins
```

## Create a plot using ggplot
 

```{r, message=FALSE}
ggplot(penguins, aes(x = species, y = flipper_length_mm)) +
  geom_boxplot() +
  labs(title = "Penguin Species vs. Flipper Length", x = "Species",
       y = "Flipper Length (mm)")

```

## Create a plot using ggpubr

Now I am going to create the same plot using ggpubr.
You will notice you do not need to use aethetics or geoms while using ggpubr. Additionally, the plot looks more "publication ready". 

```{r, message=FALSE}
ggboxplot(penguins, x = "species", y = "flipper_length_mm", 
          title = "Penguin Species vs. Flipper Length", xlab = "Species",
       ylab = "Flipper Length (mm)")
```

## ggpubr themes

# The most popular use of ggpubr is its themes. These themes make your plot "publication ready".

theme_pubr(): Create a publication ready theme

theme_pubclean(): a clean theme without axis lines, to direct more attention to the data.

labs_pubr(): Format only plot labels to a publication ready style

theme_classic2(): Create a classic theme with axis lines.

clean_theme(): Remove axis lines, ticks, texts and titles.

clean_table_theme(): Clean the the theme of a table, such as those created by ggsummarytable()


