---
title: "maddie_TPD"
author: "Maddie Wallace"
date: "2023-12-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

TPD can calculate trait probability density functions for several scales (populations, communities, species). TPD can also calculate a few functional diversity indexes. People who are interested in a functional trait-based perspective use this package. I use it mostly to visualize concepts that are sometimes hard to explain with words. Here's how you load it (it has some dependencies):
```{r}
install.packages("TPD")
install.packages("ggplot2")
install.packages("ks")
library(TPD)
library(ggplot2)
library(ks)
iris <- iris
```

Here are some fun things it can do and why I like them, using the iris data--

1. We can visualize the trait probability density of the petal length for the 3 Iris species in the dataset
```{r}
setosa_petallength <- density(iris$Petal.Length)
plot(setosa_petallength, main = "TPD petal length")
# petal lengths around 1.8 and 5 are the most likely
```


2. We can use the tpd function to calculate the trait probability densities of Iris species. Here, we will calculate the probability densities of each species considering sepal length and sepal width.
```{r}
traits_iris <- iris[, c("Sepal.Length", "Sepal.Width")]
sp_iris <- iris$Species
TPDs_iris <- TPDs(species = sp_iris, traits_iris)

# then we can visualize the TPDs in 2d (sepal length vs sepal width):
plotTPD(TPD = TPDs_iris, nRowCol = c(1,3))

# or 1d (just sepal length)
traits_iris_d1 <- iris[, c("Sepal.Length")]
sp_iris <- iris$Species
TPDs_iris_d1_a1 <- TPDs(species = sp_iris, traits = traits_iris_d1)
plotTPD(TPD = TPDs_iris_d1_a1, nRowCol = c(1,3)) 
```

We can also calculate the trait probability density of a whole plant community, to express the functional structure of a community. This uses both trait data and abundance data in the function calculations.
```{r}
abundances_comm_iris <- matrix(c(c(0.5, 0.4, 0),
                                 c(0.0, 0.9,  0.1 ),
                                 c(0.0, 0.1,  0.9 )),
                               ncol = 3, byrow = TRUE, dimnames = list(paste0("Comm.",1:3), unique(iris$Species))) 

#calculate the TPD of all species present in the community
TPDs_iris <- TPDs(species = sp_iris, traits_iris)

#calculate the community TPD
TPDc_iris <- TPDc(TPDs = TPDs_iris, sampUnit = abundances_comm_iris)

# visualize it
plotTPD(TPD = TPDc_iris, nRowCol = c(1,3))
```

The second thing I use this package for is for visualizing functional similarity/dissimilarity between species/populations/communities. There are a ton of ways to calculate functional diversity, and then you end up with a bunch of math and numbers--which is great--but I use this package when I want to visualize what the dissimilarity actually looks like (because I study trait intravariability between plant populations!)
The resulting graph is a visual of the functional dissimilarity between the three iris species sepal lengths.
```{r}

# important note-- FOR THIS CHUNK TO WORK, you have to click the green arrow in the top right corner instead of running it line by line :) no clue why. something to do with the base r plot() command and R markdown turning off something under the hood in the code chunk before the plot() can be issued

traits_iris_d1 <- iris[, c("Sepal.Length")]
sp_iris <- iris$Species
TPDs_iris_d1 <- TPDs(species = sp_iris, traits = traits_iris_d1, alpha=1)

plotTPD(TPD = TPDs_iris_d1_a1, nRowCol = c(1,3))

par(mfrow=c(1,1))
plot(TPDs_iris_d1$data$evaluation_grid, TPDs_iris_d1$TPDs[[1]], type="n",
     ylim = c(0,max(sapply(TPDs_iris_d1$TPDs, max))))

for (i in 1:length(TPDs_iris_d1$TPDs)){
  lines(TPDs_iris_d1$data$evaluation_grid, TPDs_iris_d1$TPDs[[i]], lwd=2, col=i)
}
legend("topright", bty = "n", lwd = 2, col = 1:length(TPDs_iris_d1$TPDs),
       legend = names(TPDs_iris_d1$TPDs))
```

I copy and paste this to graph other things a lot, so here's a function with all that code in it!
```{r}
tpd_plot <- function(TPDs_output) {
  par(mfrow=c(1,1))
  plot(TPDs_output$data$evaluation_grid, TPDs_output$TPDs[[1]], type="n",
       ylim = c(0,max(sapply(TPDs_output$TPDs, max))))

  for (i in 1:length(TPDs_output$TPDs)){
    lines(TPDs_output$data$evaluation_grid, TPDs_output$TPDs[[i]], lwd=2, col=i)
  }
  legend("topright", bty = "n", lwd = 2, col = 1:length(TPDs_output$TPDs),
         legend = names(TPDs_output$TPDs))

}

# see if it works!
tpd_plot(TPDs_iris_d1)

```

