---
title: 'Morpho: an R package for automated surface semi-landmarking'
author: "Alex Stewart"
date: "12/18/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(max.print = 50)
```

## Creating an Atlas

An atlas mesh essentially serves as a reference mesh, ideally with all of the traits you are interested in examining.
The positions of manually placed surface landmarks on the atlas are projected onto other meshes to automate surface landmark placement!

In Morpho, the function **createAtlas**(mesh, landmarks, patch) is used to create an atlas object for the region of interest.

1. ***mesh***: a .ply file containing a mesh of the atlas
2. ***landmarks***: a matrix containing **fixed** landmarks for the atlas and all other samples
3. ***patch***: a matrix containing surface semi-landmarks on the atlas that will be projected onto each sample in the *landmarks* matrix

The following is an example usage for createAtlas from the Morpho vignette:
```{r Creating an atlas}
library(Morpho)
data(nose)
atlas = createAtlas(shortnose.mesh, shortnose.lm[c(1:5,20:21),], shortnose.lm[-c(1:5,20:21),])
atlas
```

## Placing patches on other meshes

After an atlas has been created, the patch can be projected onto the other meshes in your sample. This is done using the **placePatch** function. The function takes three arguments and returns a matrix with the coordinates of the placed patch.

1. ***atlas***: the atlas generated using createAtlas
2. ***dat.array***: an array containing **fixed** landmarks for all samples (NOT including the atlas)
3. ***path***: a file path to a directory containing sample meshes


The following is an example usage for placePatch from the Morpho vignette:
```{r Placing surface semi-landmarks}
longnose.mesh <- tps3d(shortnose.mesh,shortnose.lm,longnose.lm,threads=1)
mesh2ply(longnose.mesh, filename="longnose")

patch_data = longnose.lm[c(1:5,20:21), ]

patched = placePatch(atlas, patch_data, prefix="longnose", path="./", inflate=5)
patched
```
Note that placePatch places things automatically, and can make mistakes that are easily visible to the human eye (e.g., placing points within a hole on a mesh). All automated placements should be checked before moving on with data processing pipelines.
