---
title: "Dslabs Ch. 21-22 Answer Key"
author: "Sabrina McNew"
date: "2023-09-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load our libraries and set plot theme
```{r, warning = FALSE, message = FALSE}
library(dslabs)
library(ggthemr)
library(tidyverse)
```

## Dslabs: Introduction to Data Science Book First edition
### Rafael Irizarry

## Chapter 21

```{r, message = FALSE}
path <- system.file("extdata", package="dslabs")
filename <- file.path(path, "fertility-two-countries-example.csv")
wide_data <- read_csv(filename)
```

### Exercises 21.5
1. Run the following command to define the co2_wide object. Use the pivot_longer function to wrangle this into a tidy dataset. Call the column with the CO2 measurements co2 and call the month column month. Call the resulting object co2_tidy.

```{r}

co2_wide <- data.frame(matrix(co2, ncol = 12, byrow = TRUE)) |> 
  setNames(1:12) |>
  mutate(year = as.character(1959:1997))
head(co2_wide) # what does this even look like 
colnames(co2_wide) # look at what we have going on 
co2_tidy <- co2_wide |> 
  pivot_longer(cols = `1`:`12`, names_to = "month", values_to = "co2") 
head(co2_tidy)
```
2. Plot CO2 versus month with a different curve for each year using this code:
```{r}
co2_tidy$month <- as.numeric(co2_tidy$month)
co2_tidy |> ggplot(aes(month, co2, color = year)) + geom_line()
str(co2_tidy)
```
3. What do we learn from this plot?
Probably b: CO2 measures are higher in the summer and have increased since 1959

4. If we think of an observation as a major, and that each observation has two variables (men admitted percentage and women admitted percentage) then this is not tidy. Use the pivot_wider function to wrangle into tidy shape: one row for each major.
```{r}
data(admissions) # the "load(admissions)" command didn't work for me
dat <- admissions |> select(-applicants)
dat |> pivot_wider(id_cols = major, names_from = gender, values_from = admitted) %>% head

```
5 - 8. Now we will try a more advanced wrangling challenge. We want to wrangle the admissions data so that for each major we have 4 observations: admitted_men, admitted_women, applicants_men and applicants_women. The trick we perform here is actually quite common: first use pivot_longer to generate an intermediate data frame and then pivot_wider to obtain the tidy data we want. We will go step by step in this and the next two exercises.

```{r}
admissions |> pivot_longer(cols = c("admitted", "applicants"), 
                           values_to = "people", names_to = "applicants")  |> 
              pivot_wider(id_cols = major, names_from = c(applicants,gender), values_from = people ) %>% head
```
### 22.4 Exercises

1. We can see column names nameFirst and nameLast. Use the left_join function to create a table of the top home run hitters. The table should have playerID, first name, last name, and number of home runs (HR). Rewrite the object top with this new table.
```{r}
#install.packages("Lahman")
library(Lahman)

top <- Batting |> 
  filter(yearID == 2016) |>
  arrange(desc(HR)) |>
  slice(1:10)

top |> as_tibble()
People |> as_tibble()

# Sabrina's notes: always good idea to know the dimensions of the dfs you're joining
dim(top)
dim(People)

# Note that top just has 10 entries. If we're left_joining to top, the resultant
# df should have 10 entries. We don't want something that has 20K entries. 
# Before saving the join to the name space, check the resulting dims. 
#
left_join(top, People) %>% dim # good news, still 10 rows, but lots more cols
top <- left_join(top, People)
```

2. Now use the Salaries data frame to add each player’s salary to the table you created in exercise 1. Note that salaries are different every year so make sure to filter for the year 2016, then use right_join. This time show first name, last name, team, HR, and salary.
```{r}
Salaries |> 
  filter(yearID == 2016) |> 
  right_join(top)  |>  
  select(nameFirst, nameLast, teamID, HR, salary) # good idea to check dims before assigning

```
3. In a previous exercise, we created a tidy version of the co2 dataset:
```{r}
co2_wide <- data.frame(matrix(co2, ncol = 12, byrow = TRUE)) |> 
  setNames(1:12) |>
  mutate(year = 1959:1997) |>
  pivot_longer(-year, names_to = "month", values_to = "co2") |>
  mutate(month = as.numeric(month))
```

We want to see if the monthly trend is changing so we are going to remove the year effects and then plot the results. We will first compute the year averages. Use the group_by and summarize to compute the average co2 for each year. Save in an object called yearly_avg.

```{r}
head(co2_wide)
yearly_avg <- co2_wide |> 
                group_by(year) |> 
                summarize(av_co2 = mean(co2, na.rm = T))
```

4. Now use the left_join function to add the yearly average to the co2_wide dataset. Then compute the residuals: observed co2 measure - yearly average.
```{r}
co2_wide |> dim() # resultant df should be 468 long
left_join(co2_wide, yearly_avg) |> dim() #good, added 1 col
co2_wide <- left_join(co2_wide, yearly_avg) |> 
  mutate(residuals = co2 - av_co2)


co2_wide %>% ggplot(aes(x = month, y = residuals, col = as.character(year))) + geom_line()
#Graph notes: at first it looked super wonky, I realized it was treating year
#as a continuous numeric variable, not a category. Adding
#as.character to year made it plot each year with a separate color 
```

